\plaintitle

### 2016-09-21 Paper/figures/figures.tex
Typo pass.

\newcommand{\anscombeFig}{
	\begin{figure*}
		\centering
		\includegraphics[width=.75\textwidth]{figures/anscombe}
		\caption{Anscombe's quartet. Each series has nearly identical summary statistics include mean, standard deviation, and linear fit. Yet, through visual inspection viewers can disambiguate  differing patterns and trends. We refer to this visual estimation of trends as ``regression by eye.''}
		\label{fig:anscombe}
	\end{figure*}
}

@@

\newcommand{\typesFig}{
	\begin{figure}
		\centering
		\includegraphics[width=.95\columnwidth]{figures/types}
		\caption{The three types of bivariate visualizations explored in this work: scatterplots, linegraphs, and area charts. The density of points made bar charts and area charts visually similar, and so we excluded bar charts from our experiments. Likewise, the error in estimating trends from heatmaps~\protect\cite{fuchs2013evaluation} made them unsuitable for the experimental task.}
		\label{fig:types}
	\end{figure}
}

>>>>

\newcommand{\anscombeFig}{
	\begin{figure*}
		\centering
		\includegraphics[width=.75\textwidth]{figures/anscombe}
		\caption{Anscombe's quartet. Each series has nearly identical summary statistics include mean, standard deviation, and linear fit. Yet, through visual inspection, viewers can disambiguate  differing patterns and trends. We refer to this visual estimation of trends as ``regression by eye.''}
		\label{fig:anscombe}
	\end{figure*}
}

@@

\newcommand{\typesFig}{
	\begin{figure}
		\centering
		\includegraphics[width=.95\columnwidth]{figures/types}
		\caption{The three types of bivariate visualizations explored in this work: scatter plots, line graphs, and area charts. The density of points made bar charts and area charts visually similar, and so we excluded bar charts from our experiments. Likewise, the error in estimating trends from heatmaps~\protect\cite{fuchs2013evaluation} made them unsuitable for the experimental task.}
		\label{fig:types}
	\end{figure}
}

### 2016-09-21 Paper/proceedings.tex
Typo pass.

Knowing the strengths and limitations of such ``estimation by eye'' is therefore important for designers of data visualizations seeking to communicate statistical quantities, especially to a general audience. On the one hand, visual estimations may be \emph{inaccurate} for the use cases intended by a designer, or they may be \emph{biased}, leading to systematic over- or under-estimations. On the other hand, while visual estimation lacks the precision of formal statistics, it may be relatively unencumbered by modeling assumptions.

@@

Ariely~\cite{ariely2001seeing} suggests that, in concert with the perception of individual objects, we also collect information about the \emph{ensemble} properties of visual displays. Szafir et al.~\cite{szafir2016four} note that this \emph{ensemble coding} might afford relatively accurate estimation of \emph{summary statistics} in visualizations. However, visualizations with good performance for summary tasks may not result in good performance at point tasks, and vice versa~\cite{albers2014task,fuchs2013evaluation}. A further difficulty is that tasks requiring estimation of values in visualizations (where there is a single correct answer) are qualitatively different from tasks requiring predictions (where differing mental models and priors can result in a multitude of potential valid responses).

@@

We analyzed data from 48 participants for each experiment (excluding rejections), for a total of 144 participants (98 male, 43 female, 3 who declined to state; $\mu_{age}= 33.2$, $\sigma_{age}=8.8$). Across the subject pool, 10 reported having graduate or professional degrees, 69 college degrees, 40 at least some college, and 25 high school diplomas. After completing the main experimental task, participants were asked to self-assess their familiarity with charts and graphs on a 5 point Likert scale. The plurality (64) rated themselves as ``3. Some familiarity,'' and none rated themselves with the maximum rating of ``5. A great deal of experience.''

@@

For each point set, we placed 100 points along a particular trend, regularly spaced in x. This both affords the use of linegraphs and area charts for displaying this data (as they require that points be one to one), and aligns well with time series data, a common domain for estimating (temporal) trends and relationships. For each point set, we created a set of residual values, sampled evenly from a Gaussian. The bandwidth of this Gaussian controls both goodness of fit and dispersion of points (see Fig. \ref{fig:sigmas}). We permuted this set of residuals, and then applied them to the original points. As heteroskedasticity introduced through permutation could alter the actual trend away from the target trend, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend. We reused these residuals across all different trend types (linear, quadratic, or trigonometric). Fig. \ref{fig:trendtypes} shows these different trend types on the same set of residuals.

@@

Our results support our first hypothesis: \textbf{larger residuals reduce accuracy at regression by eye}.  We observed a significant main effect for the bandwidth of the Gaussian used to generate residuals ($F(1,4554)=950$, $p<0.001$). The interquartile mean of absolute error increased monotonically with this bandwidth, from $0.02$ of the actual slope of the trend line when the bandwidth was $0.05$, to $0.12$ when the bandwidth was $0.20$. Figure \ref{fig:exp1sigmas} illustrates this result.

@@

%The average absolute error of slope estimation across all conditions had an interquartile mean of $0.06$ for slope values ranging from $\{-1,1\}$, corresponding to an error rate of $17\%$ of the actual slopes. 
As a point of comparison, the interquartile mean of the absolute log error, calculated as per Cleveland \& McGill~\cite{cleveland1984graphical} ($log_2 |\text{error}| + \frac{1}{8}$) was $2.4$ across all conditions. This is comparable to the error rates of many other graphical perception tasks as observed in both lab and crowdsourced studies. For instance, Heer \& Bostock~\cite{heer2010crowdsourcing}  report an absolute log error of $2.5$ for proportional judgments in treemaps, and Cleveland \& McGill~\cite{cleveland1984graphical} report an absolute log error of $2.4$ for proportional judgments in stacked bars (although compare to a log error of estimation of relative lengths of lines of $1.1$). This similarity suggests that, despite requiring estimation of aggregate statistical information, regression by eye results in judgments that are accurate enough for many practical purposes, comparable with the expected accuracy in comparing individual values in visualizations.

@@

We hypothesized that this bias would likewise occur for regression tasks when using visually asymmetric visualizations such as area charts. The slope estimation task in the previous experiment would not capture this bias, as there is no method for participants to indicate a \emph{uniform} under-estimation in trends; decreasing the slope would cause under-estimation at the beginning of the plot but not the end, and vice versa. We therefore designed this experiment to elicit estimates of the y-intercepts of trends. A within-the-bar bias would then appear as systematic under-estimation of intercept in area charts.

@@

We had one hypothesis for the second experiment:
\begin{enumerate}
	\item \textbf{Area charts would be subject to within the bar bias.} That is, participants would estimate lower values of y-intercepts of trends than with line graphs and scatter plots.
\end{enumerate}

@@

Our results support our first hypothesis. \textbf{Participant estimates were closer to the robust trend line than the trend line that included outliers}. Excluding conditions with zero outliers (and so the two trend lines would be identical), the interquartile mean of absolute error was over $4$ times higher when calculated as a comparison to the OLS trend line ($0.36$) as opposed to the robust trend line ($0.08$). A Student's T-test confirmed that the error as defined by the OLS line was significantly higher than the robust line ($t(6670)= 43$, $p<0.001$). However, a Student's T-test found a significant difference in unsigned error when compared to the robust trend line between estimates where no outliers were present, and those with any amount of outliers ($\mu=0.07$ vs. $\mu=0.08$, $t(3808) = 9.8$, $p<0.001$). Figure \ref{fig:exp3errorType} illustrates this result. This indicates that participants are not entirely ignoring outliers, although the small effect size suggests that participants are giving outliers significantly less weight than OLS regression. Figure \ref{fig:outlier} shows an example of this weighting.

Our results fail to support our second hypothesis. \textbf{While participant estimates diverge from the robust line as the number of outliers increases, the OLS trend line diverges faster than participant estimates}. There was a significant positive interaction between outlier count and absolute error with respect to the robust trend line ($R^2 = 0.02$, $F(1,4591)=110$, $p<0.001$). However, there was also a significant positive interaction between outlier count and absolute error with respect to the OLS line  ($R^2 = 0.38$, $F(1,4591)=2800$, $p<0.001$). Additionally, while the range of errors with respect to the robust line remained small (interquartile means of $0.07$ with 0 outliers, monotonically increasing to $0.10$ with 15 outliers), absolute error with respect to the OLS line increased almost 7-fold, from $0.07$ with 0 outliers, monotonically to $0.54$ with 15 outliers. These results indicate that while increased numbers of extreme values may result in participants increasing the weight of these outliers, this increase in weight is slower than the sensitivity of OLS to outliers.  Figure \ref{fig:exp3outliers} illustrates this result.

Our results also fail to support our third hypothesis, and seem to support its inverse: \textbf{There was no significant impact of outlier location on performance}. There is a confound between outlier location and OLS line: outliers on the ends of the data ranges create larger shifts in slope than outliers in the center, which may only create shifts in intercept. To evaluate our third hypothesis, we therefore defined a parameterized performance function $\frac{\text{Estimated Slope}-\text{Robust Slope}}{\text{OLS Slope} - \text{Robust Slope}}$ that contextualizes a participant estimate as an interpolation between the robust slope (value of $0$) and the OLS slope (value of $1$). A one way ANCOVA of this interpolation value as an effect of outlier location, with outlier count as a covariate, and participant ID as a random factor, found no significant effect ($F(2,3441)=0.63$, $p=0.53$).

@@

Likewise, viewers do not give the same weight to outliers as OLS regression, but they do not ignore them either. For noisy data, this robustness may be beneficial: people can be relied on to perform filtering operations without explicit guidance. In other cases, this insensitivity may be undesirable. For instance, more complicated automatic analytical processes may use OLS regression or other outlier sensitive techniques as part of their calculations, resulting in a disconnect between human and statistical conclusions at later stages of the sensemaking process. Similarly, the human tendency to discount or downweight outliers may result in slow adaptation to new data that does not fit into existing patterns.

@@

Our experimental setup was intentionally simple, affording only a single free parameter for each experiment. In actual regression by eye, the viewer may simultaneously engage in multiple types of estimation: choosing a particular type of fit, ignoring outliers, and estimating the parameters of the chosen model. Errors in any one of these steps could compound, resulting in performance worse than our measures, where many of these decisions are fixed. It is also likely that there are many visual features, such as the convex hull of points, or the numerosity of point clouds, that are acting as visual proxies for estimation of the trend. Given the unimodal error function and uniform density of our experimental stimuli, these proxies are useful for estimating the trend line. However, in datasets where internal densities of points may be skewed, or the envelope of point clouds non-informative for estimating central tendency, these visual proxies may introduce biases and confounds in visual estimations.

@@

In this paper, we examine the ability of visualization viewers to perform \emph{regression by eye}: the visual estimation of trends in bivariate visualizations. We show that viewers without statistical training can reliably estimate both linear and non-linear trends in charts such as scatter plots and line graphs. However, area graphs are subject to a ``within the bar'' bias, leading to estimated trends with lower intercepts than other bivariate visualizations. Regression by eye is also less sensitive to outliers than standard least-squares regression. This decreased sensitivity results in a divergence between trends estimated by viewers, which do not give much weight to extreme outliers, and those trends calculated by regression methods that can be heavily influenced by a few extreme values.

>>>>

Knowing the strengths and limitations of such ``estimation by eye'' is therefore important for designers of data visualizations seeking to communicate statistical quantities, especially to a general audience. On the one hand, visual estimations may be too \emph{inaccurate} for the use cases intended by a designer, or they may be \emph{biased}, leading to systematic over- or under-estimations. On the other hand, while visual estimation lacks the precision of formal statistics, it may be relatively unencumbered by modeling assumptions.

@@

Ariely~\cite{ariely2001seeing} suggests that, in concert with the perception of individual objects, we also collect information about the \emph{ensemble} properties of visual displays. Szafir et al.~\cite{szafir2016four} note that this \emph{ensemble coding} might afford relatively accurate estimation of \emph{summary statistics} in visualizations. However, visualizations with good performance for summary tasks may not result in good performance at point tasks, and vice versa~\cite{albers2014task,fuchs2013evaluation}. A further difficulty is that tasks requiring estimation of values in visualizations (where there is a single correct answer) are qualitatively different from tasks requiring predictions (where differing mental models and priors can result in a multitude of potentially valid responses).

@@

We analyzed data from 48 participants for each experiment (excluding rejections), for a total of 144 participants (98 male, 43 female, 3 who declined to state; $M_{age}= 33.2$, $SD_{age}=8.8$). Across the subject pool, 10 reported having graduate or professional degrees, 69 college degrees, 40 at least some college, and 25 high school diplomas. After completing the main experimental task, participants were asked to self-assess their familiarity with charts and graphs on a 5 point Likert scale. The plurality (64) rated themselves as ``3. Some familiarity,'' and none rated themselves with the maximum rating of ``5. A great deal of experience.''

@@

For each point set, we placed 100 points along a particular trend, regularly spaced in x. This both affords the use of line graphs and area charts for displaying this data (as they require that points be one to one), and aligns well with time series data, a common domain for estimating (temporal) trends and relationships. For each point set, we created a set of residual values, sampled evenly from a Gaussian. The bandwidth of this Gaussian controls both goodness of fit and dispersion of points (see Fig. \ref{fig:sigmas}). We permuted this set of residuals, and then applied them to the original points. As heteroskedasticity introduced through permutation could alter the trend away from the target trend, we performed rejection sampling to ensure that the slope of the trend of the resulting points was within $10^{-7}$ of the target. We reused these residuals across all different trend types (linear, quadratic, or trigonometric). Fig. \ref{fig:trendtypes} shows these different trend types on the same set of residuals.

@@

Our results support our first hypothesis: \textbf{larger residuals result in less accuracy at regression by eye}.  We observed a significant main effect for the bandwidth of the Gaussian used to generate residuals ($F(1,4554)=950$, $p<0.001$). The interquartile mean of absolute error increased monotonically with this bandwidth, from $0.02$ of the actual slope of the trend line when the bandwidth was $0.05$, to $0.12$ when the bandwidth was $0.20$. Figure \ref{fig:exp1sigmas} illustrates this result.

@@

%The average absolute error of slope estimation across all conditions had an interquartile mean of $0.06$ for slope values ranging from $\{-1,1\}$, corresponding to an error rate of $17\%$ of the actual slopes. 
Prior work in graphical perception often measures the absolute log error: e.g., Cleveland \& McGill~\cite{cleveland1984graphical} calculate their performance metric as ($log_2 |\text{error}| + \frac{1}{8}$). As a point of comparison the interquartile mean of the absolute log error, calculated as per  was $2.4$ across all conditions in this experiment. Differences in methodology and measurement discourage statistical inferences comparing this value with those in similar experiments. Nevertheless, error rates for regression by eye are at least in the same magnitude as those observed in other graphical perception tasks in both lab and crowdsourced studies. For instance, Heer \& Bostock~\cite{heer2010crowdsourcing} report an absolute log error of $2.5$ for proportional judgments in treemaps, and Cleveland \& McGill~\cite{cleveland1984graphical} report an absolute log error of $2.4$ for proportional judgments in stacked bars (although compare to a log error of estimation of relative lengths of lines of $1.1$). This similarity suggests that, despite requiring estimation of aggregate statistical information, regression by eye results in judgments that are accurate enough for many practical purposes, comparable with the expected accuracy in comparing individual values in visualizations.

@@

We hypothesized that this bias would likewise occur in regression by eye when using visually asymmetric visualizations such as area charts. The slope estimation task in the previous experiment would not capture this bias, as there is no method for participants to indicate a \emph{uniform} under-estimation in trends; decreasing the slope would cause under-estimation at the beginning of the plot but not the end, and vice versa. We therefore designed this experiment to elicit estimates of the y-intercepts of trends. A within-the-bar bias would then appear as systematic under-estimation of intercept in area charts.

@@

We had one hypothesis for the second experiment:
\begin{enumerate}
	\item \textbf{Area charts would be subject to within the bar bias.} That is, participants would estimate lower values of y-intercepts of trends in area charts, as opposed to line graphs and scatter plots.
\end{enumerate}

@@

Our results support our first hypothesis. \textbf{Participant estimates were closer to the robust trend line than the trend line that included outliers}. Excluding conditions with zero outliers (and so the two trend lines would be identical), the interquartile mean of absolute error was over $4$ times higher when calculated as a comparison to the OLS trend line ($0.36$) as opposed to when calculated from the robust trend line ($0.08$). A Student's T-test confirmed that the error as defined by the OLS line was significantly higher than the robust line ($t(6670)= 43$, $p<0.001$). However, a Student's T-test found a significant difference in unsigned error when compared to the robust trend line between estimates where no outliers were present, and those with any amount of outliers ($\mu=0.07$ vs. $\mu=0.08$, $t(3808) = 9.8$, $p<0.001$). Figure \ref{fig:exp3errorType} illustrates this result. This indicates that participants are not entirely ignoring outliers, although the small effect size suggests that participants are giving outliers significantly less weight than OLS regression. Figure \ref{fig:outlier} shows an comparison of robust, participant, and OLS trend lines.

Our results fail to support our second hypothesis. \textbf{While participant estimates diverge from the robust line as the number of outliers increases, the OLS trend line diverges faster than participant estimates}. There was a significant positive interaction between outlier count and absolute error with respect to the robust trend line ($R^2 = 0.02$, $F(1,4591)=110$, $p<0.001$). However, there was also a significant positive interaction between outlier count and absolute error with respect to the OLS line  ($R^2 = 0.38$, $F(1,4591)=2800$, $p<0.001$). Additionally, while the range of errors with respect to the robust line remained small (interquartile means of $0.07$ with 0 outliers, monotonically increasing to $0.10$ with 15 outliers), absolute error with respect to the OLS line monotonically increased almost 7-fold, from $0.07$ with 0 outliers to $0.54$ with 15 outliers. These results indicate that, while increased numbers of extreme values may result in participants increasing the weight of these outliers, this increase in weight is slower than the sensitivity of OLS to outliers.  Figure \ref{fig:exp3outliers} illustrates this result.

Our results also fail to support our third hypothesis: \textbf{There was no significant impact of outlier location on performance}. There is a confound between outlier location and OLS line: outliers on the ends of the data ranges create larger shifts in slope than outliers in the center, which may only create shifts in intercept. To evaluate our third hypothesis, we therefore defined a parameterized performance function $\frac{\text{Estimated Slope}-\text{Robust Slope}}{\text{OLS Slope} - \text{Robust Slope}}$ that contextualizes a participant estimate as an interpolation between the robust slope (value of $0$) and the OLS slope (value of $1$). A one way ANCOVA of this interpolation value as an effect of outlier location, with outlier count as a covariate, and participant ID as a random factor, found no significant effect of outlier location on interpolative value ($F(2,3441)=0.63$, $p=0.53$).

@@

Likewise, while viewers do not give the same weight to outliers as OLS regression, they do not ignore them either. For noisy data, this robustness may be beneficial: people can be relied on to perform filtering operations without explicit guidance. In other cases, this insensitivity may be undesirable. For instance, more complicated automatic analytical processes may use OLS regression or other outlier sensitive techniques as part of their calculations, resulting in a disconnect between human and statistical conclusions at later stages of the sensemaking process. Similarly, the human tendency to discount or downweight outliers may result in slow adaptation to new data that does not fit into existing patterns.

@@

Our experimental setup was intentionally simple, affording only a single free parameter for each experiment. In actual regression by eye, the viewer may simultaneously engage in multiple types of estimation: choosing a particular type of fit, ignoring outliers, and estimating the parameters of the chosen model. Errors in any one of these steps could compound, resulting in performance worse than our measures, where many of these decisions are fixed \emph{a priori}. It is also likely that there are many visual features, such as the convex hull of points, or the numerosity of point clouds, that are acting as visual proxies for estimation of the trend. Given the unimodal error function and uniform density of our experimental stimuli, these proxies are useful for estimating the trend line. However, in datasets where internal densities of points may be skewed, or the envelope of point clouds non-informative for estimating central tendency, these visual proxies may introduce biases and confounds in visual estimations.

@@

In this paper, we examine the ability of visualization viewers to perform \emph{regression by eye}: the visual estimation of trends in bivariate visualizations. We show that viewers without statistical training can reliably estimate both linear and non-linear trends in charts such as scatter plots and line graphs. However, area graphs are subject to a ``within the bar'' bias, leading to estimated trends with lower intercepts than other bivariate visualizations. Regression by eye is also less sensitive to outliers than standard least-squares regression. This decreased sensitivity results in a divergence between trends estimated by viewers, which do not give much weight to extreme outliers, and those trends calculated by regression methods, that can be heavily influenced by a few extreme values.
